### 2020-06-03 writeup/transfer.tex
revisions

It is well-known since the seminal work of Caruana \cite{C97} that how well multi-task learning performs depends on task relatedness.
We formalize this connection in the following simplified setting.

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Recall that each task has sample size $n_i = \rho_i \cdot p$, for $i = 1, 2$.
	And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denotes the covariates of the two tasks, respectively.
	We assume that for the target task, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma_i^2$, for $i=1,2$.
	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{app_proof_sec3}.

@@

\begin{proposition}[Task model distance]\label{prop_dist_transition}
	In the isotropic model, suppose that $\rho_1 > 1$ and $\sigma_1 = \sigma_2 = \sigma$.
	Then we have that
	%Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between $\Psi(\beta_1, \beta_2)$ and $\Phi(\rho_1, \rho_2)$:
	\squishlist
		\item If $\Psi(\beta_1, \beta_2) < \gamma_{+}^{-1}  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Psi(\beta_1, \beta_2) \ge \gamma_{-}^{-1}  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})$.
	\squishend
	Here $\gamma_{-} = (1-\oo(1))\cdot(1 - \rho_1^{-1/2})^4$ and $\gamma_{+} = (1+\oo(1))\cdot(1 + \rho_1^{-1/2})^4$.
	Concretely, if $\rho_1 > 40$, then $\gamma_{-} \in (1,2)$ and $\gamma_{+} \in (1/2, 1)$.
\end{proposition}

Proposition \ref{prop_dist_transition} simplifies Theorem \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
Prosposition \ref{prop_dist_transition} explains the transition observed in Figure \ref{fig_model_shift}.
Note that there are several unexplained observations near the transition threshold $0$, which are caused by the concentration errors $\gamma_+, \gamma_-$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}. %, which is based on our main result described later in Theorem \ref{thm_model_shift}.

\textbf{Algorithmic consequence.}
We can in fact extend the result to the cases where the noise variance is different.
In this case, we will see that MTL is particularly effective.
%As an extension of Proposition \ref{prop_dist_transition}, we observe that MTL is particular powerful when the labeled data of the source task is less noisy compared to the target task.
Concretely, suppose the noise variance $\sigma_1^2$ of task $1$ differs from the noise variance $\sigma_2^2$ of task $2$.
If $\sigma_1^2$ is too large, the source task provides a negative transfer to the target.
If $\sigma_1^2$ is small, the source task is more helpful.
We leave the result to the Appendix in Proposition \ref{prop_var_transition}.
Inspired by the observation, we propose a single-task based metric to help understand MTL results using STL results.
\squishlist
	\item For each task, we train a single-task model.
	Let $z_s$ and $z_t$ be the prediction accuracy of each task, respectively.
	Let $\tau\in(0, 1)$ be a fixed threshold.
	\item If $z_s - z_t > \tau$, then we predict that there will be positive transfer when combining the two tasks using MTL.
	If $z_s - z_t < -\tau$, then we predict negative transfer.
\squishend

@@

\begin{proposition}[Source/target sample size]\label{prop_data_size}
	In the isotropic model, assume that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
	We have that
	\squishlist
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\gamma_+} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \gamma_+^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > \frac{1}{\gamma_-} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1.5) - \gamma_{-}^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{proposition}
Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the data ratio $n_1 / n_2$.
From Figure \ref{fig_size}, we see that our result is able to explain the transition from positive to negative transfer.
There are several unexplained observations near $y = 0$ caused by the errors $\gamma_-, \gamma_+$.
The proof of Proposition \ref{prop_data_size} can be found in Appendix \ref{app_proof_32}.

\textbf{Connection to Taskonomy.} We use our tools to explain a key result of Taskonomy by Zamir et al.'18 \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $R \cdot n_i$ datapoints from every task.
The data efficiency ratio roughly scales as
\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
equal to \[ \frac{1}{\rho_1 + \rho_2} {+ \frac{1}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$, the transfer is positive.
Hence the data efficiency ratio quantifies how effective is the positive transfer using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.
The precise statement can be found in Appendix \ref{app_proof_32}.

@@

\begin{example}\label{ex_complement}
	We measure covariate shift by the matrix $M = \Sigma_1^{1/2} \Sigma_2^{-1/2}$.
	Assume that $\Psi(\beta_1, \beta_2) = 0$ for simplicity.
	We compare two cases: (i) when $M = \id$; (ii) when $M$ has $p/2$ singular values that are equal to $\lambda$ and $p/2$ singular values that are equal to $1 / \lambda$.
	Hence, $\lambda$ measures the severity of covariate shift.
	Figure \ref{fig_covariate} shows a simulation of this setting by varying $\lambda$.
	We observe that as $n_1/n_2$ increases, the performance gap between the case of $M = \id$ and $\lambda = 2$ or $4$ increases for MTL.
\end{example}

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
To compare different choices of $M$ on the performance of $\hat{\beta}_t^{\MTL}$, let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M$ in decreasing order.
Let $\mu_{\min} < \mu < \mu_{\max}$ are fixed values that do not grow with $p$.
Consider the following bounded set
\begin{align*}
		\cS_{\mu}\define\bigset{M \mid \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_p, \lambda_1 \le \mu_{\max}},
\end{align*}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.

\begin{proposition}[Covariate shift]\label{prop_covariate}
	Assume that $\Psi(\beta_1, \beta_2) = 0$ and $\rho_1, \rho_2>1$.
	Let $g(M)$ denote the test error of $\hat{\beta}_t^{\MTL}$ when the covariance shift matrix is equal to $M\in\cS_{\mu}$.
	We have that \[ g(\mu\id) \le \bigbrace{1+ \bigo{\frac{\rho_2}{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
The following proposition shows that when $n_1 \gg n_2$, $\te(\hat{\beta}^{\MTL})$ is minimized when there is no covariate shift between source and target task.
The proof of Proposition \ref{prop_covariate} is left to Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@

\textbf{Algorithmic consequence.}
Our observation highlights the need to correct covariate shift when $n_1 / n_2$ is large.
Hence for such settings, we expect procedures that aim at correcting covariate shift to provide more significant gains.
We consider a covariance alignment procedure proposed in Wu et al.'20 \cite{WZR20}, which is designed for the purpose of correcting covariate shift.
The idea is to add an alignment module between the input and the shared module $B$.
This new module is then trained together with $B$ and the output layers.
We validate our insight on this procedure in the experiments.

>>>>

It is well-known since the seminal work of Caruana \cite{C97} that how well multi-task learning performs depends on task relatedness. We formalize this connection in the following simplified setting, where we can perform explicit calculations.

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Recall that each task has sample size $n_i = \rho_i \cdot p$, for $i = 1, 2$.
	And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	We assume that for the target task, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma_i^2$, for $i=1,2$.
	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

@@

\begin{proposition}[Task model distance]\label{prop_dist_transition}
	In the isotropic model, suppose that $\rho_1 > 1$ and $\sigma_1 = \sigma_2 = \sigma$.
	Then we have that
	%Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between $\Psi(\beta_1, \beta_2)$ and $\Phi(\rho_1, \rho_2)$:
	\squishlist
		\item If $\Psi(\beta_1, \beta_2) < \gamma_{+}^{-1}  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Psi(\beta_1, \beta_2) > \gamma_{-}^{-1}  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
	Here $\gamma_{-} = (1-\oo(1))\cdot(1 - \rho_1^{-1/2})^4$ and $\gamma_{+} = (1+\oo(1))\cdot(1 + \rho_1^{-1/2})^4$.
	Concretely, if $\rho_1 > 40$, then $\gamma_{+} \in (1,2)$ and $\gamma_{-} \in (1/2, 1)$.
\end{proposition}

Proposition \ref{prop_dist_transition} simplifies Corollary \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
Prosposition \ref{prop_dist_transition} explains the transition observed in Figure \ref{fig_model_shift}.
Note that there are several unexplained observations near the transition threshold $0$, which are caused by the concentration errors $\gamma_+$ and $\gamma_-$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}. %, which is based on our main result described later in Theorem \ref{thm_model_shift}.

\textbf{Algorithmic consequence.}
We can in fact extend the result to the cases where the noise variances are different.
In this case, we will see that MTL is particularly effective.
%As an extension of Proposition \ref{prop_dist_transition}, we observe that MTL is particular powerful when the labeled data of the source task is less noisy compared to the target task.
Concretely, suppose the noise variance $\sigma_1^2$ of task $1$ differs from the noise variance $\sigma_2^2$ of task $2$.
If $\sigma_1^2$ is too large, the source task provides a negative transfer to the target.
If $\sigma_1^2$ is small, the source task is more helpful.
We leave the result to Proposition \ref{prop_var_transition} in Appendix \ref{app_proof_31}.
Inspired by the observation, we propose a single-task based metric to help understand MTL results using STL results.
\squishlist
	\item For each task, we train a single-task model.
	Let $z_s$ and $z_t$ be the prediction accuracy of each task, respectively.
	Let $\tau\in(0, 1)$ be a fixed threshold.
	\item If $z_s - z_t > \tau$, then we predict that there will be positive transfer when combining the two tasks using MTL.
	If $z_s - z_t < -\tau$, then we predict negative transfer.
\squishend

@@

\begin{proposition}[Source/target sample size]\label{prop_data_size}
	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
	Then we have that
	\squishlist
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\gamma_+} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \gamma_+^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > \frac{1}{\gamma_-} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1.5) - \gamma_{-}^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{proposition}
Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the data ratio $n_1 / n_2$.
From Figure \ref{fig_size}, we see that our result is able to explain the transition from positive to negative transfer.
There are several unexplained observations near $y = 0$ caused by the errors $\gamma_-, \gamma_+$.
The proof of Proposition \ref{prop_data_size} can be found in Appendix \ref{app_proof_32}.

\textbf{Connection to Taskonomy.} We use our tools to explain a key result of Taskonomy by Zamir et al. (2018) \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $x \cdot n_i$ datapoints from every task. The data efficiency ratio is defined as
\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
roughly 
\[ \frac{1}{\rho_1 + \rho_2} {+ \frac{2}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$ (up to a constant multiple), the transfer is positive.
Moreover, the data efficiency ratio quantifies how effective the positive transfer is using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.
The precise statement can be found in Proposition \ref{prop_data_efficiency} in Appendix \ref{app_proof_32}.

@@

\begin{example}\label{ex_complement}
	We measure covariate shift by the matrix $M = \Sigma_1^{1/2} \Sigma_2^{-1/2}$.
	Assume that $\Psi(\beta_1, \beta_2) = 0$ for simplicity.
	We compare two cases: (i) when $M = \id$; (ii) when $M$ has $p/2$ singular values that are equal to $\lambda$ and $p/2$ singular values that are equal to $1 / \lambda$.
	Hence, $\lambda$ measures the severity of the covariate shift.
	Figure \ref{fig_covariate} shows a simulation of this setting by varying $\lambda$.
	We observe that as $n_1/n_2$ increases, the performance gap between case (i) $M = \id$ and case (ii) with $\lambda = 2$ or $4$ increases for MTL.
\end{example}

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
To compare different choices of $M$ regarding the performance of $\hat{\beta}_t^{\MTL}$, let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M$ in decreasing order.
Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.
Consider the following bounded set
\begin{align*}
		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_p\le \lambda_1 \le \mu_{\max}\right.},
\end{align*}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.

\begin{proposition}[Covariate shift]\label{prop_covariate}
	Assume that $\Psi(\beta_1, \beta_2) = 0$ and $\rho_1, \rho_2>1$.
	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when the covariance shift matrix is equal to $M\in\cS_{\mu}$.
	We have that \[ g(\mu\id) \le \bigbrace{1+ \bigo{\frac{\rho_2}{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
This proposition shows that when $n_1 \gg n_2$, $\te(\hat{\beta}^{\MTL})$ is minimized when there is no covariate shift between source and target tasks.
The proof of Proposition \ref{prop_covariate} is left to Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@

\textbf{Algorithmic consequence.}
Our observation highlights the need to correct covariate shift when $n_1 / n_2$ is large.
Hence for such settings, we expect procedures that aim at correcting covariate shift to provide more significant gains.
We consider a covariance alignment procedure proposed in Wu et al. (2020) \cite{WZR20}, which is designed for the purpose of correcting covariate shift.
The idea is to add an alignment module between the input and the shared module $B$.
This new module is then trained together with $B$ and the output layers.
We validate our insight on this procedure in the experiments.
